---
title: "Getting started with ZICDT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ZICDT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`ZICDT` learns a **directed tree (or forest)** in which each node is an entire
**zero-inflated composition** on the probability simplex. It scores candidate
edges with a Kullback--Leibler divergence (finite on exact zeros), models each
child's conditional mean as a mixture of a baseline and a parent-driven,
column-stochastic transition map, fits edge parameters by EM (in C++), searches
for the globally optimal spanning arborescence with the Chu--Liu/Edmonds
algorithm, and selects the edge penalty by cross-validation.

```{r}
library(ZICDT)
```

## Simulate data on a known tree

Each node is an `n x d_j` matrix whose rows are compositions. Here we build a
5-node tree with edges `1->2->3` and `1->4->5`.

```{r}
G <- matrix(0, 5, 5)
G[1, 2] <- 1; G[2, 3] <- 1; G[1, 4] <- 1; G[4, 5] <- 1

sim <- zicdt_simulate(G, d = c(8, 10, 12, 9, 11), n = 100, seed = 7)
lapply(sim$data, dim)          # one n x d_j matrix per node
```

## Fit at a fixed penalty

`zicdt_fit()` learns the globally optimal structure for a given penalty `alpha`.
Note the parameterization: `alpha` multiplies the KL fit, so **larger `alpha`
yields a denser structure**.

```{r}
fit <- zicdt_fit(sim$data, alpha = 2)
fit
zicdt_edges(fit)
```

## Choose the penalty by cross-validation (main workflow)

`zicdt_cv()` runs K-fold cross-validation over a grid of penalties, selects the
one minimizing out-of-sample KL loss, and refits the final tree. This is the
usual entry point.

```{r}
cvfit <- zicdt_cv(sim$data, alpha_grid = c(seq(0.1, 1, 0.1), 2, 5), K = 3)
cvfit
cvfit$alpha        # selected penalty
head(cvfit$cv)     # per-penalty CV table
```

The learned directed tree is available as an adjacency matrix
(`cvfit$adjacency`, with `adjacency[k, j] == 1` meaning edge `k -> j`), as a tidy
edge list via `zicdt_edges(cvfit)`, and as an `igraph` object (`cvfit$graph`).
The fitted edge transition matrices are in `cvfit$params`.
